Nos\'e-Hoover Chains Improved by Backward Error Analysis: Application to Rigid-Body Molecular Dynamics

### 2018-02-11 rigid_bodies.tex
New section with proposed method at the beginning

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville

@@

\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}

@@

Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman2010} involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field. However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski2012, Patra2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton1995} has been applied to simulate small molecules,\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}

The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables. The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible. Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble. The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

@@

\section{Theory}
\label{sec:theory}

Guided by a particular factorization of the rotation matrix expressed in terms of quaternion components, in a previous paper \cite{Silveira_2017} we derived symplectic numerical integrators for rigid bodies in the NVE ensemble. The first one uses an exact solution for torque-free rotations. The second approach is based on the method introduced by \citeauthor{Dullweber_1997} \cite{Dullweber_1997} and \citeauthor{Miller2002} \cite{Miller2002}, that employs an approximate solution for rotation. For mathematical details the reader is referred to our paper.\cite{Silveira_2017} Next, we restrict ourselves to present the expressions required to develop the NVT case.

@@

When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled. The position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$. In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$ and, as a consequence, $\tr{\vt q}{\dot{\vt q}} = 0$.

The matrix $\mt A$ which can be computed from a given unit quaternion $\vt q$ as explained in Refs.~\citenum{Allen1989,Miller2002} admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C},
\end{equation} 
where
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
That factorization of matrix $\mt A$ provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components $\dot{\vt q}$, which is given by
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.

In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}. In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a diagonal matrix in the principal frame. For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Silveira_2017}

>>>>

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\shadow}[1]{\widetilde{#1}}

@@

%\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}
\title{Nos\'e-Hoover Chains Improved by Backward Error Analysis: Application to Rigid-Body Molecular Dynamics}

@@

Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman_2010} involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field. However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski_2012, Patra_2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj_2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton_1995} has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}

The Nos\'{e}-Hoover chain thermostat,\cite{Martyna_1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables. The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible. Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble. The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj_2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

@@

\section{Proposed Method}
\label{sec:methodology}

In a previous paper \cite{Silveira_2017}, we employed a particular factorization of the orientation matrix expressed in terms of quaternion components to 1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for rigid bodies in the NVE ensemble and 2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} as a sequence of uniform rotations around the principal axes of a body, thus unveiling the equivalence between this method and an older one by \citeauthor{Dullweber_1997} \cite{Dullweber_1997}

The ordinary differential equation (ODE) system that describes a Hamiltonian rigid-body motion are \cite{Silveira_2017}
\begin{subequations}
	\label{eq:ODE system for NVE}
	\begin{align}
	&\dot{\vt r} =
	%+\diff{\Ham{}}{\vt p} =
	{\mt M}^{-1} {\vt p}, \\
	&\dot{\vt p} =
	%-\diff{\Ham{}}{\vt r} = 
	{\vt F}, \\
	&\dot{\vt q} =
	%+\diff{\Ham{}}{\vt \pi} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
	&\dot{\vt \pi} =
	%-\diff{\Ham{}}{\vt q} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
	\end{align}
\end{subequations}
with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$. In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines the body's orientation, and $\vt \pi$ is the momentum conjugated to such quaternion. Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity. Matrices $\mt M$ and $\mt I$ are diagonal ones, with the former containing the body's mass in all its diagonal entries and the latter containing the principal moments of inertia. Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the principal-axis frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the principal-axis and space-fixed frames) expressed as a function of the quaternion components. The fact that such matrix admits the simple factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ played a central role in Ref.~\citenum{Silveira_2017}.

The Hamiltonian conserved by Eq.~\eqref{eq:ODE system for NVE} is
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the body's kinetic and potential energies, respectively. While $U(\vt r, \vt q)$ can have any functional form, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

An advantage of this notation is that Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the respective properties of all bodies piled together. The same occurs for diagonal matrices $\mt M$ and $\mt I$, whose sizes increase to $3N \times 3N$ in this case. In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns. Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged to accommodate a number of free atoms.

A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system's configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$. A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$. In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, where the asterisk denotes the state of the system immediately before a propagation takes place. The action of $e^{\timestep \Liu{A}}$ is a uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) with simultaneous torque-free rotation during a full time step.

Description of analytical and NO\_SQUISH rotations...

Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble. For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement. To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$. The equations of motion, in this case, are
\begin{subequations}
	\label{eq:ODE system for NVT}
	\begin{align}
	&\dot{\vt r} =
	{\mt M}^{-1} {\vt p}, \\
	&\dot{\vt p} =
	{\vt F} - \alpha_1 {\vt p}, \\
	&\dot{\vt q} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \\
	&\dot{\vt \pi} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
	&\dot{\eta}_j = \alpha_j, \text{ and} \\
	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M,
	\end{align}
\end{subequations}

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where each $Q_j$ is an inertial parameter of the corresponding thermostat. Besides, each $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1,
\end{align}
\end{subequations}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set as null, as usual (otherwise, $L = 6N$).\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
%Finally, $\shadow K$ is the total kinetic energy of the system, whose computation (not using Eq.~\ref{eq:kinetic energy}, hence the distinct symbol) is the central feature of the method we propose.

The ``extended energy'' which is conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H_\text{\tiny NVT} = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

As in the previous case, it is possible to devise a splitting method to solve Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}. By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation*}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation*}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain. Of course, the NVE part can be integrated exactly as described before. The NHC part, in turn, can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

In our scheme, the first propagator to act, $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$. Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied sequentially with $j$ ranging from $M-1$ down to $1$. With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as
\begin{align*}
&p_{\eta_j} = p_{\eta_j}^\ast + \Big( G_j^\ast - \alpha_{j+1}^\ast p_{\eta_j}^\ast \Big) \phi\left(\frac{\alpha_{j+1}^\ast \timestep}{4n}\right) \frac{\timestep}{4n} \; \text{and} \\
&\eta_{j+1} = \eta_{j+1}^\ast + \alpha_{j+1}^\ast \frac{\timestep}{4n}.
\end{align*}

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ enacts the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of the thermostat coordinate $\eta_1$. Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\shadow K$ to be scaled by the square of it, hence ${\shadow K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\shadow K}^\ast$. In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\shadow K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_1}$ to $e^{\frac{\timestep}{4n} \mathcal{L}_M}$ are applied again, but now in the increasing order of their indices. It is worth noting that the $\eta$'s are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H_\text{\tiny NVT}$.

The whole procedure described above is repeated $n$ times. Because of its low computational cost, the compensation required for increasing the time step $\timestep$ without degrading the accuracy of the NHC part (that is, employing a larger $n$) causes a small impact in the overall integration effort. In contrast, evaluating the NVE part is expensive for involving force/torque computations and its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}. A high-order integration scheme\cite{Omelyan_2007, vanZon_2008} could possibly admit larger time steps, but it would raise both cost and complexity because for entailing the computation (or numerical estimation) of force/torque gradients. The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.

Here we take a simpler path. Instead of attempting to improve the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we try to quantify the obtained deviations and then use them to tune the thermostatting procedure and properly weight the sampled configurations when canonical averages are computed.

A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are also Hamiltonian. Thus, a trajectory generated to be roughly consistent with the specified Hamiltonian $\Ham{}$ will be exactly consistent with a nearby, albeit unknown, function $\shadow{\mathcal H}$. It is also known that $\shadow{\mathcal H}$, referred to as a shadow Hamiltonian, depends on the time step size $\timestep$ and that, for a time-reversible splitting like Eq.~\eqref{eq:splitting NVE}, $\shadow{\mathcal H} = \Ham{} + \mathcal{O}(\timestep^2)$. Estimating the trailing term of this equality, either analytically or numerically, is a task of backward error analysis. As demonstrated in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be better approximated as $\shadow{\Ham{}} = \shadow K + \shadow U + \mathcal{O}(h^4)$, where $\shadow K$ and $\shadow U$ will be called shadow kinetic and shadow potential energies, respectively. The former is given by
\begin{equation}
\label{eq:shadow kinetic energy}
\shadow K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\shadow{\mathbf \Omega}$ is a complicated matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, but independent of both $\vt p$ and $\vt \pi$. The shadow potential is also independent of these momenta and is given by
\begin{equation}
\label{eq:shadow potential energy}
\shadow U = U - \frac{\timestep^2}{24} \left[ \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A}(\vt q) {\mt I}^{-1} {\mt A}(\vt q) {\vt \tau} \right].
\end{equation}

Therefore, the outcome of solving Eq.~\eqref{eq:ODE system for NVE} using a Trotter-Suzuki splitting will be close to the exact solution of Eq.~\eqref{eq:ODE system for NVE}, but it will be even closer to the solution of a system composed of equations
\begin{equation}
\label{eq:shadow ODE system}
\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] =
\left[\begin{array}{c} \diff{\shadow K}{\vt p} \\ \diff{\shadow K}{\vt \pi} \end{array}\right] =
\shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
as well as $\dot{\vt p} = -\diff{\shadow{\mathcal H}}{\vt r}$ and $\dot{\vt \pi} = -\diff{\shadow{\mathcal H}}{\vt q}$. In our proposed method, we acknowledge this fact and thus use the Nos\'e-Hoover chain to provide a canonical distribution consistent with $\shadow{\mathcal H}$ (rather than $\Ham{}$) at temperature $T$. Note that this entails solving Eq.~\eqref{eq:ODE system for NVT} exactly as described before, with the only requirement of replacing $K$ by $\shadow K$ in the definition of the generalized force $G_1$. By contrasting Eqs.~\eqref{eq:generalized force NVT}, \eqref{eq:shadow kinetic energy}, and \eqref{eq:shadow ODE system}, we can see that
\begin{equation}
G_1 = 2 \shadow K - L k_\text{\tiny B} T.
\end{equation}

Direct evaluation of Eq.~\eqref{eq:shadow kinetic energy} for computing $\shadow K$ is quite complex and expensive. However, one can do it very easily using the approach devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the derivatives in the left-hand side of Eq.~\eqref{eq:shadow ODE system}. Denoting the obtained estimators as $\dot{\vt r}_\text{\tiny NVE}$ and $\dot{\vt q}_\text{\tiny NVE}$ and substituting Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:shadow kinetic energy}, we find out that
\begin{equation}
\shadow K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}_\text{\tiny NVE} + \tr{\vt \pi} \dot{\vt q}_\text{\tiny NVE} \right).
\end{equation}

- Pratical estimation of numerical derivatives: virtual steps and interpolation

- Scaling of $\shadow K$ during the NHC integration.

Reweighting:
\begin{equation}
\label{eq:reweighting}
\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
\end{equation}

\section{Theoretical Background}
\label{sec:theory}

@@

When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of a center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t) = \tr{[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}. In this way, the position of any constituent particle $j$ will vary with time as $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in a body-fixed frame of reference centered in $\vt r$ and ${\mt A}(\vt q)$ is the body's orientation matrix (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components. The superscript $t$ stands for transposition, which is equivalent to inversion in the case of a canonical transformation like $\mt A$ \cite{Goldstein_2002}. The central feature explored in Ref.~\citenum{Silveira_2017} is the fact that $\mt A$ admits the simple factorization
\begin{equation}
\label{eq:factorization_of_A}
{\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q),
\end{equation}
where
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

This factorization creates an intermediate frame of reference for representing three-dimensional vectors in $\mathbb{R}^3$ as quaternions that lie in the hyperplane orthogonal to $\vt q$, since the columns of matrix $\mt B$ and those of matrix $\mt C$ form two different bases for such a subspace of $\mathbb{R}^4$. Transformations amongst three distinct reference frames are enacted by multiplications with these matrices or their transposes, such as represented in the following scheme:
\[
\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
\]
